---
title: "R for physical scientists"
author: "Matt Lacey and Taha Ahmed"
output: 
  revealjs::revealjs_presentation:
    css: theme.css
    center: true
    transition: slide
    highlight: tango
---
  
```{r, echo = F}
library(knitr)
library(magrittr)
```

# 2: Data manipulation and visualisation

## Packages we'll use today!

`ggplot2`, `dplyr`, `tidyr`

Let's load them:

```{r}
library(tidyverse)
```

## `dplyr`

- A fast, consistent tool for working with data frame like objects, both in memory and out of memory.
- Identify the most important data manipulation tools needed for data analysis and make them easy to use from R.
- Provide blazing fast performance for in-memory data by writing key pieces in C++.
- Use the same interface to work with data no matter where it's stored, whether in a data frame, a data table or database.

## `tidyr`

- Provides functions for "tidying" data - `gather()` and `spread()`

# Introduction to `dplyr`

##

`dplyr` provides several functions for manipulating data frames, e.g.,

`select()`, `filter()`, `mutate()`, `rename()`

##

Remember the `mtcars` dataset from yesterday?

```{r}
str(mtcars)
```

##

`select()` selects specific columns from a data frame:

```{r}
mtcars2 <- select(mtcars, mpg, cyl, disp)
str(mtcars2)
```

##

`select()` can also be used to drop specific columns, with `-`:

```{r}
mtcars3 <- select(mtcars, -am, -gear, -carb)
str(mtcars3)
```

##

`rename()` renames specified columns. Use `rename(data, newname = oldname)`. You can rename as many columns as you want in one `rename()` call.

```{r}
mtcars4 <- rename(mtcars, ncyl = cyl, weight = wt)
str(mtcars4)
```

##

`filter()` selects rows matching given conditions.

```{r}
mtcars5 <- filter(mtcars, cyl == 6, am == 1)
mtcars5
```

##

Multiple conditions can be specified for the same variable.

```{r}
mtcars6 <- filter(mtcars, hp > 100, hp < 200)
str(mtcars6)
```

## Comparison operators

- `x < y` - x is less than y
- `x > y` - x is greater than y
- `x <= y` - x is less than or equal to y
- `x >= y` - x is greater than or equal to y
- `x == y` - x is equal to y
- `x != y` - x is not equal to y

## Value matching

For matching values to a vector, use `%in%`:

```{r}
mtcars7 <- filter(mtcars, cyl %in% c(6, 8))
str(mtcars7)
```

##

`mutate()` adds new columns, preserving all previous ones.

```{r}
mtcars8 <- mutate(mtcars, displ_l = disp / 61.0237)
str(mtcars8)
```

##

Note: all `dplyr` methods ignore row names in data frames (on purpose). If you have them and want to keep them, they have to be converted to an explicit variable. `dplyr` has a function for this, `rownames_to_column`:

```{r}
has_rownames(mtcars) # check if rownames exist
mtcars9 <- rownames_to_column(mtcars, "name")
str(mtcars9)
```

## Exercise!

Look at the `iris` dataset (e.g., `View(iris)`).

Use `filter()` to make a new data frame `df`, for only the "setosa" species with a sepal length between 5 and 6.5 cm.

## Solution!

Your code should look something like this:

```{r}
df <- filter(iris, Species == "setosa", Sepal.Length >= 5, Sepal.Length <= 6.5)
head(df)
```

##

# Let's practice `ggplot()`...

## The `diamonds` dataset

There is a built-in dataset in the `ggplot2` packages called `diamonds`.

```{r}
str(diamonds)
```

(Try also `?diamonds` to find out more.)

##

```{r, fig.width = 6, fig.height = 4}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

##

Use `aes()` to map properties to variables:

```{r, fig.width = 6, fig.height = 4}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) # <<----
```

##

Properties not mapped to variables should not be inside `aes()`

```{r, fig.width = 6, fig.height = 4}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6) # <<----
```

## variable `<-` plot

```{r, fig.width = 6, fig.height = 4}
p <- ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6) # <<----

print(p)
```

## adding layers

```{r, fig.width = 6, fig.height = 4}
p <- p + ggtitle("diamonds")

print(p)
```

## Scaling

There are a very large number of functions for modifying properties mapped to variables, such as `x`, `y`, `size`, `shape`, `alpha`, `color`, `fill`, etc.

They all begin with `scale_`...

##

Scaling axes can be done with, for example, `scale_x_continuous()` and `scale_y_continuous()`, which removes values outside the range...

```{r, fig.width = 6, fig.height = 4, warning = F}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6) +
  scale_x_continuous(limits = c(0, 3)) + # <<----
  scale_y_continuous(limits = c(0, 10000)) # <<----
```

##

another alternative is to use or `coord_cartesian()`, which effectively rescales the plot window - this can be useful sometimes:

```{r, fig.width = 6, fig.height = 4}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6) +
  coord_cartesian(xlim = c(0, 3), ylim = c(0, 10000)) # <<----
```

##

There are useful functions for changing colour schemes based on specially suited colour palettes.

- `scale_color_brewer()` and `scale_fill_brewer()` for **discrete** data
- `scale_color_distiller() and `scale_fill_distiller()` for **continuous** data

##

```{r, fig.width = 6, fig.height = 4, warning = F}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6) +
  scale_color_brewer(palette = "Set1") # <<----
```

##

```{r, fig.width = 6, fig.height = 4, warning = F}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = x * y), size = 2, alpha = 0.6) +
  scale_color_distiller(palette = "YlGnBu", limits = c(10, 120)) # <<----
```

##

... and many more possibilities

- `scale_x_log10()`
- `scale_size_discrete()`
- `scale_fill_continuous()`
- `scale_alpha_manual()`

and so on...

I recommend reading the `ggplot2` documentation to learn more!

## Labelling

Use `xlab()` and `ylab()` to label axes:

```{r, fig.width = 6, fig.height = 4, warning = F}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6) +
  scale_x_continuous(limits = c(0, 3)) + 
  scale_y_continuous(limits = c(0, 10000)) +
  xlab("weight / carats") + # <<----
  ylab("price / USD") # <<----
```

## Facetting

```{r, fig.width = 6, fig.height = 4}
diamonds2 <- filter(diamonds, cut %in% c("Premium", "Ideal"), 
                    clarity %in% c("VVS1", "VVS2", "IF"))

p2 <- ggplot(diamonds2, aes(x = carat, y = price)) +
  geom_point(aes(color = color), size = 2, alpha = 0.5)

print(p2)
```

##

Create panels for each value of a variable with `facet_grid(rows ~ columns)`

```{r, fig.width = 6, fig.height = 4}
p2 + facet_grid(. ~ cut)
```

##

```{r, fig.width = 7, fig.height = 4.8}
p2 + facet_grid(clarity ~ cut)
```

##

Use `facet_wrap()` if you have a 1-dimensional sequence of panels and want to wrap it into a fixed number of rows or columns:

```{r, fig.width = 7, fig.height = 4.8}
p2 + facet_wrap( ~ color, ncol = 4)
```


# `dplyr -> ggplot`

## `%>%`

`%>%` is the "pipe" operator.

It comes from the `magrittr` package and is loaded automatically along with `dplyr`.

##

`%>%` "pipes" an object to the first argument of a function, i.e:

```{r, eval = F}
x %>% f(y, z)
```

is the same as:

```{r, eval = F}
f(x, y, z)
```

This creates code which is easily read left-to-right.

##

For example:

```{r, fig.width = 6, fig.height = 4}
diamonds %>%
  ggplot(aes(x = carat, y = price)) +
    geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6)
```

##

```{r, fig.width = 6, fig.height = 4}
filter(diamonds, cut %in% c("Premium", "Ideal")) %>%
  ggplot(aes(x = carat, y = price)) +
    geom_point(aes(color = cut), size = 2, shape = 21, alpha = 0.6)
```

##

In cases where you don't want the object on the left hand side to be the first argument in the function, use the dot (`.`) as placeholder:

```{r, eval = F}
y %>% f(x, ., z)
```

is equivalent to `f(x, y, z)`

# More advanced data manipulation

##

`group_by` is a `dplyr` function that converts a data frame into a grouped table, with groups defined by variables in the dataset.

```{r}
grouped_iris <- iris %>%
    group_by(Species)
str(grouped_iris)
```