---
title: "R for physical scientists - suggestions for topics"
author: "Matt Lacey and Taha Ahmed"
output: 
  revealjs::revealjs_presentation:
    css: theme.css
    center: true
    transition: slide
    highlight: tango
---

## Basics

- What is R
- What are its strengths
    - Made for importing data, doing analysis and making graphs 
    - Extensibility and package system
    - Community support (StackOverflow, etc)
- The environment (R, RStudio)
- `?`
   
## Interpreter

Some examples, like:

```{r}
2 + 2
```

## Objects

- `<-`
- Vectors, matrices, data frames etc
    - `str` and `class` functions
- Indexing
    - `[]`, `$` etc

## Functions

- Some important ones, like `c`, `head`, `tail`, etc
- Some little hints at what R is actually doing, like showing how

```{r}
2 + 3
```

is the same as

```{r}
`+`(2, 3)
```

- How to write a simple function which takes multiple arguments

## `apply` functions

- At least `sapply` and `lapply`
- Can mention `for` loops but `lapply` should be the first option. `while` also useful

## ggplot

How to make a simple plot

```{r, eval = F}
ggplot()
```

How to add data

```{r, eval = F}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point()
```

How to tweak it however you like it

```{r, eval = F}
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(aes(color = cyl), size = 2) +
  facet_wrap( ~ cyl) + 
  xlab ...
  ...
```

## Transforming data on plots

```{r, echo = F}
library(ggplot2)
```

```{r, fig.height = 3, fig.width = 5}
ggplot(mtcars, aes(x = disp, y = 282.5/mpg)) +
  geom_point(aes(color = factor(cyl))) +
  ylab("l/100km")
```

## Importing data

- `read.table`, `read.delim`, `read.csv`
- `readxl` package

## Manipulating data

- `dplyr`: `filter` and `select` especially

## Other important packages

- `reshape2` (`melt` and `cast`)
- `lubridate`
- `magrittr`
- `scales` and `grid` for ggplots

and possibly some others but these are the ones I use and am mostly familiar with. Could mention that packages exist for mix in code from C++, Python, MATLAB, etc.

## Reporting

- Sweave and knitr (probably the latter mostly)
- R Markdown
- R Notebooks!!
- Shiny??